---
title: "Graphical EDA I: continous data"
author: Team 1
date: "`r format(Sys.time(), '%d %B %Y')`"
output: slidy_presentation
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE)
```

## Structure

**1. Examining Continuous Variables**

**2. Looking for Structure: Dependency Relationships and Associations**

**3. Investigating Multivariate Continuous Data**

*3.1. Introduction*

*3.2. What is a parallel coordinate plot (pcp)?*

*3.3. Features you can see with parallel coordinate plots*

*3.4. Interpreting clustering results*

*3.5. Parallel coordinate plots and time series*

*3.6. Parallel coordinate plots for indices*

*3.7. Options for parallel coordinate plots*

*3.8. Main points*


## 1. Examining Continuous Variables

## 2. Looking for Structure: Dependency Relationships and Associations

## 3. Investigating Multivariate Continuous Data

### **3.1. Introduction**

#### Traditional methods used for displaying multivariate continous data:
- **scatterplots**. *Disadvantage:* only for bivariate data, not effective for exploring in higher dimensions;
- **matrices of scatterplots**. *Disadvantage:* do not really convey what multivariate structure there might be in the data;
- **dimension reduction methods (PCA, MDS, etc).** *Disadvantage:* they approximate the data, and it is difficult to assess how good the approximations are.

#### **New Method**: parallel coordinate plots. 
#### This solution has become popular for multivariate continuous data in recent years.

## **3.2. What is a parallel coordinate plot (pcp)?**

```{r fig.width=10,fig.height=5}
library('GGally'); data(food, package="MMST")
names(food) <- c("Fat", "Food.energy", "Carbohyd", "Protein", "Cholest", "Wt", "Satur.Fat")
# food1 <- food/food$Wt
food1 <- head(food, 10)/head(food, 10)$Wt
# TODO: sterge linia de deasupra si decomenteaza pe cealalta
# TODO: pt un plot mai plin si mai mult timp de asteptat la build
ggparcoord(data = food1, columns=c(1:5, 7), scale="uniminmax", alphaLines=0.2) + 
  xlab("") + ylab("")
```

#### In a parallel coordinate plot:
- all axes are parallel to one another;
- each variable has its own individual vertical axis (or alternatively all the axes are horizontal);
- each axis is usually scaled from the minimum to the maximum case values for the variable;
- each polygonal line across all axes defines a ***profile***.

\

#### **Functions for drawing pcp's in R**

| function       | package |
|----------------|---------|
| *ggparcoord*   | GGally  |
| *parcoord*     | MASS    |
| *cparcoord*    | gclus   |
| *pcp*          | PairViz |
| *parallelplot* | lattice |
| *parallel.ade* | epade   |
| *ipcp*         | iplots  |
| *...*          | ...     |

## **3.3. Features you can see with parallel coordinate plots**

#### Parallel coordinate plots are very useful for checking results like:
- outliers
- groups of cases with very similar values (**clusters**)
- bivariate associations between adjacent variables (**correlations**)
- gaps or concentrations of data
- skewness of univariate distributions for several variables at once

\

#### **Illustrating Fisher’s iris dataset**
```{r fig.width=10,fig.height=5}
ggparcoord(iris, columns=1:4, groupColumn="Species")
```

## **3.4. Interpreting clustering results**

#### We can compare the clusters graphically with a pcp in which the clusters are given different colours.

#### In the **USArrests** dataset from 1973, the *Assault* variable separates the clusters, because it has a larger range of values than the other three variables.

```{r fig.width=10,fig.height=5}
hcav <- hclust(dist(USArrests), method="ave")
clu3 <- cutree(hcav, k=3)
clus <- factor(clu3)
usa1 <- cbind(USArrests, clus)
ggparcoord(usa1, columns=1:4, groupColumn="clus", scale="uniminmax") + 
  xlab("") + ylab("") + 
  theme(legend.position = "none") +
  geom_line(size = 1.1)
ggparcoord(usa1, columns=1:4, groupColumn="clus", scale="globalminmax") + 
  xlab("") + ylab("") + 
  theme(legend.position = "none") +
  geom_line(size = 1.1)
```

## **3.5. Parallel coordinate plots and time series**

#### In a parallel coordinate plot used for representing time series:
- each axis represents one of the time points;
- we cannot change the order of the axis;
- a common scaling is used for all axes (**scale**=*globalminmax*).

**Note:** *ipcp* method of **iplots** package is better for creating interactive time series pcp's.

\

#### **Illustrating the acres of corn planted by US states from 1866 to 2011**
```{r fig.width=10,fig.height=5}
library(reshape2); data(nass.corn, package="agridat")
c1 <- melt(nass.corn, id=c("year", "state"))
c1 <- within(c1, StateV <- interaction(state, variable))
c2 <- dcast(c1, StateV~year)
ggparcoord(subset(c2[1:48,], c2[1:48,147]> 250000), columns=2:147, groupColumn="StateV", scale="globalminmax") + 
  xlab("Year") + ylab("Acres") + 
  scale_x_discrete(breaks=seq(1865, 2015, 10)) + 
  theme(legend.position = "none")
```

## **3.6. Parallel coordinate plots for indices**

#### Usually, index values are weighted combinations of the values of their components. Examples:
- **stock market indices** - used to represent performances of groups of shares;
- **consumer price index** - summarises prices across a wide range of products;
- **average teaching score** - an overall index used for ranking universities in UK.

#### One way to represent them in pcp's:
- the index itself will have one axes associated;
- for each of the components there will be a seperate axes;
- the polygonal lines are the profiles of the individuals;
- more effective if some individuals or groups are identified by colour.

\

#### **Illustrating the Russell group from the *uniranks* dataset**
```{r fig.width=10,fig.height=5}
data(uniranks, package="GDAdata")
names(uniranks)[c(5, 6, 8, 10, 11, 13)] <- c("AvTeach", "NSSTeach", "SpendperSt", "Careers", "VAddScore", "NSSFeedb")
uniranks1 <- within(uniranks, StaffStu <- 1/(StudentStaffRatio))
uniranks2 <- within(uniranks1,
                    Rus <- ifelse(UniGroup=="Russell", "Russell", "not"))
ggparcoord(uniranks2[order(uniranks2$Rus, decreasing=FALSE),],
           columns=c(5:8, 10:14),
           order=c(5,12,8,9,14,6,13,7,11,10),
           groupColumn="Rus", scale="uniminmax") +
           xlab("") + ylab("") +
           theme(legend.position = "none",
           axis.ticks.y = element_blank(),
           axis.text.y = element_blank()) +
           scale_colour_manual(values = c("gray","red")) + 
           geom_line(size = 0.8)
```

# THANK YOU!
